---
title: "NYPD Shooting Incident Data Report"
author: "T.Panzarino"
date: "2024-02-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in NYPD Data

1.  Assign NYPD data URL to variable "url"
2.  Read in data using read_csv()

```{r get NYPD Data}
library(tidyverse)
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
NYPD_data = read_csv(url)

```

## Clean up data

1.  Replace (null) and UNKNOWN values with NA to create a consistent way to filter data
2.  Replace "Dwelling" with "Housing" as they contain the same types of locations
3.  Remove columns that contain information not useful to us, such as coordinates and jurisdiction code. We still have Borough information, Location (Indoor/Outdoor), precinct, location classification and location description to describe where the shooting took place. Jurisdiction is not necessary if we have precinct and borough information.
4.  Rename columns to shorter, cleaner versions.

```{r clean NYPD data}

NYPD_data[NYPD_data == "(null)"] <- NA
NYPD_data[NYPD_data == "NONE"] <- NA
NYPD_data[NYPD_data == "UNKNOWN"] <- NA
NYPD_data[NYPD_data == "DWELLING"] <- "HOUSING"

library(lubridate)

NYPD_data <- NYPD_data %>%
  select(-c(Latitude, Longitude, X_COORD_CD, Y_COORD_CD, Lon_Lat, JURISDICTION_CODE)) %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%
  rename(
    date = OCCUR_DATE,
    time = OCCUR_TIME,
    borough = BORO,
    indoor_outdoor = LOC_OF_OCCUR_DESC,
    loc_categ = LOCATION_DESC,
    loc_desc = LOC_CLASSFCTN_DESC,
    murder = STATISTICAL_MURDER_FLAG,
    perp_age= PERP_AGE_GROUP,
    perp_sex = PERP_SEX,
    prep_race = PERP_RACE,
    vic_age = VIC_AGE_GROUP,
    vic_sex = VIC_SEX,
    vic_race =  VIC_RACE,
    precinct = PRECINCT,
    incident = INCIDENT_KEY
  ) 
```

4.  View data summary

```{r original data summary}
summary(NYPD_data)

```

## Create new data sets from NYPD data

1.  Create data set for total shootings/ day using count().

```{r NYPD totals}
NYPD_totals <- NYPD_data %>%
  count(date) %>%
  rename(incidents = n)
  

```

```{r totals summary}

summary(NYPD_totals)

```

2.  Create data set for total shootings/ day per borough using pivot_wider and count(). Rename the boroughs to lowercase form.

```{r NYPD borough totals}
NYPD_borough_totals <- NYPD_data %>%
  count(date, borough) %>%
  pivot_wider(names_from = borough, values_from = n) %>%
  rename(
    brooklyn = BROOKLYN,
    queens = QUEENS,
    manhattan = MANHATTAN,
    staten_island = `STATEN ISLAND`,
    bronx = BRONX
  )
  
```

```{r boroughs summary}

summary(NYPD_borough_totals)

```
3. Create data set with total number of incidents per location description. I removed other and NA because it was too much of an outlier to plot against other locations later on.

```{r locations breakdown}
NYPD_locs <- NYPD_data %>%
  count(loc_desc) %>%
  rename(
    incidents = n)

NYPD_locs <-NYPD_locs[-c(3, 9), ]


```


##Visualizations

1.  Create plot for all incidents

```{r total indicents}
NYPD_totals %>%
  filter(incidents >= 0 ) %>%
  ggplot(aes(x = date, y = incidents)) +
  geom_line(aes(color = "incidents")) +
  #geom_point(aes(color = "n")) +
  #geom_line(aes(y = BROOKLYN, color = "BROOKLYN")) +
  #geom_point(aes(y = BROOKLYN, color = "BROOKLYN")) +
  #scale_y_log10() +
  ylim(-10,70) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = "NYPD: Total Shooting Incidents", y = NULL)
```

2.  Create plot for boroughs. Due to the volume of data, only Bronx, Brooklyn and Queens were plotted between November 2022 and end of year 2022.

```{r total indicents/ borough}
NYPD_borough_totals %>%
  
  ggplot(aes(x = date, y = bronx)) +
  geom_line(aes(color = "Bronx")) +
  geom_point(aes(color = "Bronx")) +
  geom_line(aes(y = brooklyn, color = "Brooklyn")) +
  geom_point(aes(y = brooklyn, color = "Brooklyn")) +
  geom_line(aes(y = queens, color = "Queens")) +
  geom_point(aes(y = queens, color = "Queens")) +
  #geom_line(aes(y = manhattan, color = "Manhattan")) +
  #geom_point(aes(y = manhattan, color = "Manhattan")) +
  #geom_line(aes(y = staten_island, color = "Staten Island")) +
  #geom_point(aes(y = staten_island, color = "Staten Island")) +
  ylim(-5,15) +
  xlim(as.Date(c('2022-11-01', '2023-01-01'), format="%Y-%m-%d"))+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = "NYPD: Shooting Incidents in Bronx, Brooklyn, Queens from November 2022 through the End of Year", y = NULL)
```
 
 3. Create plot for locations
 
```{r location plot}
 ggplot(NYPD_locs, aes(x = loc_desc, y = incidents)) +
 geom_bar(stat = "identity") +
 theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
 labs(title = "NYPD: Total Shooting Incidents Based on Location, 2006-2022", y = NULL)

```

##Model

1.Use a linear model to see if there is any relationship between date and number of shootings
2. Determine difference between predicted numbers and actual numbers
3. Plot difference bewteen predicted and actual values


```{r first data model}
mod <- lm(incidents ~ date, data = NYPD_totals)
summary(mod)

totals_w_pred <- NYPD_totals %>%
  mutate(pred = predict(mod))

summary(totals_w_pred)
```

```{r model plot}

totals_w_pred %>% ggplot() +
  geom_point(aes(x = date, y = incidents), color = "blue") +
  geom_point(aes(x = date, y = pred), color = "red")

```

## Conclusion/Bias

Having grown up on Long Island, I am very familiar with the NYC boroughs and this may have increased the amount of Bias I have towards this data. I automatically know which boroughs have a more suburban layout versus the more urban ones, leading me to assume there would be more shootings in the urban locations. I also know which areas have historically experienced more violence. I believe that the location data is the most biased, as there appear to be many different descriptions that may or may not be describing the same type of location.  I think that this data set should have included population size of each borough, so that shootings could be explored in proportion to population size. This may be a more accurate way to portray this data than volume of incidents alone. Based on the date plotting/model, it appears that there may be a slight trend of shootings going down over time in the 5 boroughs. It appears that during the holiday season (November 01- end of year 2022) there are less shootings and less variations in shooting amounts in queens compared to the Bronx and Brooklyn. After creating a box plot of total number of shootings per locations, it appears most shootings happen on the street, followed by housing. 



